\documentclass[11pt]{letter}
\usepackage[margin=1in]{geometry}
\usepackage{times}

\signature{Aaron W. Storey\\
Department of Computer Science\\
Clarkson University\\
Potsdam, NY 13699\\
\texttt{storeyaw@clarkson.edu}\\
ORCID: 0009-0009-5560-0015}

\address{Aaron W. Storey\\
Department of Computer Science\\
Clarkson University\\
Potsdam, NY 13699}

\begin{document}
\begin{letter}{Editor-in-Chief\\
IEEE Transactions on Artificial Intelligence}

\opening{Dear Editor,}

I am pleased to submit the manuscript entitled ``\textbf{World Weaver: Cognitive Memory Architecture for Persistent World Models in Agentic AI Systems}'' for consideration for publication in IEEE Transactions on Artificial Intelligence.

\textbf{Summary:} This paper presents a tripartite cognitive memory architecture (episodic, semantic, procedural) that enables AI agents to build persistent, inspectable world models. The work addresses a fundamental limitation of current large language models: their inability to retain knowledge across sessions.

\textbf{Key Contributions:}
\begin{enumerate}
    \item A cognitive memory architecture grounded in Tulving's memory taxonomy and Anderson's ACT-R framework
    \item Hybrid retrieval combining dense semantic vectors with sparse lexical matching, achieving 84\% Recall@10 vs. 72\% for dense-only search ($p < 0.001$)
    \item Systematic survey of 52 papers on AI agent memory (2020--2024)
    \item Critical analysis of limitations, failure modes, and fundamental open questions
\end{enumerate}

\textbf{Significance:} As AI systems become more capable and consequential, the question of how they accumulate and organize knowledge becomes critical for alignment, efficiency, and safety. This work provides both a practical architecture and a conceptual framework for addressing the ``amnesia problem'' in agentic AI.

\textbf{Declarations:}
\begin{itemize}
    \item No conflicts of interest to declare
    \item This manuscript has not been submitted elsewhere
    \item Code and data will be released upon acceptance at: \texttt{https://github.com/astoreyai/world-weaver}
\end{itemize}

Thank you for considering this submission. I look forward to the opportunity to contribute to IEEE Transactions on Artificial Intelligence.

\closing{Sincerely,}

\end{letter}
\end{document}
